---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Reproducible Research: Peer Assessment 1
==========================================

## Loading and preprocessing the data
### Read data from activity.csv
```{r readdata}
data=read.csv("activity.csv")
```
### Convert date from factor to date
```{r converdate}
data$date=as.Date(data$date)
```

## What is mean total number of steps taken per day?
### Calculate number of steps per day
```{r calcnoofsteps}
perdaydata=aggregate(steps~date,data=data,sum)
head(perdaydata)
```
### Create histogram of the total number of steps per day
```{r histofnoofsteps}
hist(perdaydata$steps,xlab="steps",main = "Histogram of the total number of steps per day",ylim = c(0,40))
```

### Calculate mean
```{r meanofsteps,cache=TRUE}
stepmean=mean(perdaydata$steps)
# round and store only upto 2 decimal places
stepmean=format(round(stepmean, 2), nsmall = 2)
stepmedian=median(perdaydata$steps)
```

The mean of total number of steps per day is `r stepmean`.  
The median of total number of steps per day is `r stepmedian`.  

## What is the average daily activity pattern?

### Calculate average step data per slot over all days
```{r meanstepperslot}
meanperslotdata=aggregate(steps~interval,data=data,FUN="mean")
head(meanperslotdata)
```

### Draw Plot

```{r plotmeanstepsperslot}
plot(meanperslotdata$steps~meanperslotdata$interval,type="l",xlab="Intervals",ylab="Steps",main="Average no. of steps per interval",col="blue")
```


###Calculate interval that, on average across all the days in the dataset, contains the maximum number of steps.
```{r maxstepinslots}
maxstepinterval=meanperslotdata[meanperslotdata$steps==max(meanperslotdata$steps),]$interval
```

That interval is the `r maxstepinterval`th interval.  

## Imputing missing values

### Calculate number of missing values in the data
```{r calcmissingcount}
missingvalnum=nrow(data)-nrow(data[complete.cases(data),])
```
Number of missing values in the data are `r missingvalnum`.  

### Create new data with missing values replaced by average number of steps per interval accross all days

We are creating  replacing the missing values witht the average number of steps per interval accross all days  


```{r createnewdata}
newdata<-data.frame()
for(i in 1:nrow(data)){
        if(is.na(data[i,]$steps))
        {
                step<-meanperslotdata[meanperslotdata$interval==data[i,]$interval,]$steps
                newdata<-rbind(newdata,c(step, data[i,]$date,data[i,]$interval))
                
                
        }else
        {
                
                newdata<-rbind(newdata,c(data[i,]$steps,data[i,]$date,data[i,]$interval))
              
        }
      colnames(newdata)  = colnames(data)
     
      
}
newdata$date=as.Date(newdata$date,origin="1970-01-01")
```

### Calculate number of steps per day for new data which has no missing value
```{r calcnoofstepsnewdata}
newperdaydata=aggregate(steps~date,data=newdata,sum)
head(newperdaydata)
```
### Create histogram of the total number of steps per day for new data which has no missing value
```{r histofnoofstepsnewdata}
hist(newperdaydata$steps,xlab="steps",main = "Histogram of the total number of steps per day",ylim = c(0,40))
```

### Calculate mean for new data which has no missing value
```{r meanofnoofstepsnewdata}
newmean=mean(newperdaydata$steps)
# round and store only upto 2 decimal places
newmean=format(round(newmean, 2), nsmall = 2)
newmedian=median(newperdaydata$steps)
# round and store only upto 2 decimal places
newmedian=format(round(newmedian, 2), nsmall = 2)

```

The mean of total number of steps per day for new data is `r newmean`.  
The median of total number of steps per day for new data is `r newmedian`.  

### Calculate diff in new and old mean and median
```{r calcdiffinmeanmedian}
diffmean=(as.numeric(stepmean)-as.numeric(newmean))
diffmedian=(as.numeric(stepmedian)-as.numeric(newmedian))

```

The diff in old and new mean is `r diffmean`  

The diff in old and new median is `r diffmedian`  


## Are there differences in activity patterns between weekdays and weekends?

### Create factor variable for weekday
```{r weekdayvariable}
newdata$weekday=""
for(i in 1:nrow(newdata)){
        if(weekdays(newdata[i,]$date) %in% c("Saturday","Sunday"))
        {
                newdata[i,]$weekday="Weekend"
        }
        else
                newdata[i,]$weekday="Weekday"
                
}

newdata$weekday=factor(newdata$weekday)

head(newdata)
```


### Calculate average steps per interval accross all Weekdays

```{r meanstepsallweekday}

meanperslotweekdaydata=aggregate(steps~interval,data=newdata[newdata$weekday=="Weekday",],FUN="mean")
head(meanperslotweekdaydata)

```

### Calculate average steps per interval accross all Weekends

```{r meanstepsallweekend}

meanperslotweekenddata=aggregate(steps~interval,data=newdata[newdata$weekday=="Weekend",],FUN="mean")
head(meanperslotweekenddata)

```

### Create plots for average steps per interval accross all Weekdays, Weekends
```{r plotmeanstepsweekdayweekend,fig.height=8}
par(mfrow=c(2,1))
plot(meanperslotweekdaydata$steps~meanperslotweekdaydata$interval,type="l",xlab="Intervals",ylab="Steps",main="Average no. of steps per interval across all weekdays",ylim=c(-2,250),col="blue")
plot(meanperslotweekenddata$steps~meanperslotweekenddata$interval,type="l",xlab="Intervals",ylab="Steps",main="Average no. of steps per interval across all weekends",ylim=c(-2,250),col="blue")
```
